---
title: Stan and BRMS introduction
author: Fiona Seaton
date: 2020-07-02
output: html_document

draft: false
toc: false
type: docs

linktitle: Stan and brms
menu:
  misc:
    weight: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This tutorial offers a short overview of using the brms package to fit Bayesian regression models in Stan. This tutorial is also available online on my [website](https://www.fionamseaton.com/tutorial/misc/brms-examples/).

## stan overview

Stan is a platform used for Bayesian modelling. Unlike JAGS and BUGS the underlying MCMC algorithm is Hamiltonian - meaning it uses gradients rather than steps. Stan uses a variant of a No-U-Turn Sampler (NUTS) to explore the target parameter space and return the model output. 

In practice, this means:

* Better at exploring the model space
* More likely to find issues with the model parameterisation
* Quicker than JAGS/BUGS with more complex models
* LOADS of [diagnostics](https://mc-stan.org/misc/warnings.html)

Stan can be interfaced to from various software, the most commonly used and well supported is R but there are also options to interface from python or the command line. Within R there is the rstan package which does the direct interfacing with stan (along with StanHeaders), but there are also many helper packages for fitting stan models including rstanarm and brms.

```{r stan software ecosystem, echo = FALSE, message = FALSE}
library(ggdag)
library(dplyr)
library(ggplot2)
dagify(rstan ~ stan,
       pystan ~ stan,
       cmdstan ~ stan,
       rstanarm ~ rstan,
       brms ~ rstan) %>% 
  ggdag_classic(layout = "tree") + theme_dag()
```

There are also several other packages in R that work with stan models, such as bayesplot, loo, shinystan etc. 

Both rstanarm and brms use formula notation in the style of lme4 in order to specify stan models. The main difference in between the two packages is that rstanarm has all of their models pre-specified and compiled into stan code while brms writes and compiles a new stan model each time. This means rstanarm can be a lot quicker than brms, but brms supports a wider range of model types. I use brms exclusively as I am a creature of habit and learnt it first, so that is what I will present here.

# Installation
A guide to installing rstan can be found online [here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started), it is now much easier than it used to be - just install off CRAN as standard. It will take a few minutes, and afterwards you need to check whether your C++ toolchain is correctly set up using pkgbuild. Their github page also gives an optional step to configure the toolchain.

```{r rstan install, eval = FALSE}
install.packages("rstan")

# check toolchain
pkgbuild::has_build_tools(debug = TRUE)

# Optional - configure toolchain
dotR <- file.path(Sys.getenv("HOME"), ".R")
if (!file.exists(dotR)) dir.create(dotR)
M <- file.path(dotR, ifelse(.Platform$OS.type == "windows", "Makevars.win", "Makevars"))
if (!file.exists(M)) file.create(M)
cat("\nCXX14FLAGS=-O3 -march=native -mtune=native",
    if( grepl("^darwin", R.version$os)) "CXX14FLAGS += -arch x86_64 -ftemplate-depth-256" else 
    if (.Platform$OS.type == "windows") "CXX11FLAGS=-O3 -march=corei7 -mtune=corei7" else
    "CXX14FLAGS += -fPIC",
    file = M, sep = "\n", append = TRUE)
```

Other packages that rely on rstan can be installed from CRAN/github as usual, I won't go into the details here.


```{r package load, message = FALSE}
library(brms)
library(dplyr)
library(ggplot2)
theme_set(theme_classic())
```


# Simple model example

I'm going to start by running a very simple mixed model here in order to demonstrate how easy fitting a model with brms can be. All the data here is from the agridat package, which is a package that holds several agricultural related datasets.

```{r get and plot data}
library(agridat)
dat <- ilri.sheep
ggplot(dat, aes(x = gen, y = weanwt)) +
  geom_boxplot()
```


The brms model for this (with default priors, i.e. this is not a recommended workflow!):
```{r simple brms model}
mod1 <- brm(weanwt ~ gen - 1 + (1|ewe) + (1|ram), data = ilri.sheep, cores = 4)
```
```{r brms simple model plots}
summary(mod1)
plot(mod1, ask = FALSE, N = 4)
```


# Example suggested workflow
When using these methods it is suggested that you think more about the prior assumptions that you are putting into the model. Several people within the stan community are now advocating a model building approach that follows several steps. I'm going to give a quick outline of the kind of steps that I follow when building models.

First, prior predictive checks. Here we take the model structure and priors we are suggesting and evaluate the data structure that is implied by these priors.

Binomial data - Phytophtera disease occurrence in a pepper field.

```{r}
dat <- gumpertz.pepper %>%
  mutate(disease = recode(disease, "Y" = 1, "N" = 0))
ggplot(dat, aes(x = leaf, y = disease, colour = water)) + geom_jitter()
```

First, define the model and find out what priors are automatically given by brms.

```{r}
get_prior(y ~ trt - 1 + (1|block), data = beall.webworms,
          family = poisson)
```

So we can see that the default is to have a student T prior on the intercept and random effect. Let's put a wide prior on b

```{r}
pr <- prior(normal(0,10), class = "b")
```


Now check by sampling the prior what kind of data this suggests:
```{r}
mod_pr <- brm(y ~ trt - 1 + (1|block), data = beall.webworms,
              family = poisson, prior = pr,
              cores = 4, sample_prior = "only")
```

There is handy function within stan that allows you to see what the data suggested by the model looks like - pp_check. I will discuss this more later when we come to posterior checks but the default plots the density of the data and the model predicted data. You can also use plot(conditional_effects(<mod>)) to plot the predicted effects of the treatments.

```{r}
pp_check(mod_pr)
plot(conditional_effects(mod_pr))
```

These plots show that our prior suggests that having counts of millions/billions is a possible outcome, which both seems unreasonable and could lead to issues with model convergence as the model fitting process has to explore these unlikely regions of model space. We can try this with tighter priors and see if it makes the model more sensible.

```{r}
pr <- prior(normal(0,.5), class = "b") +
  prior(student_t(3,0,1), class = "sd") +
  prior(student_t(3,0,1), class = "Intercept")
```



Now check by sampling the prior what kind of data this suggests
```{r}
mod_pr <- update(mod_pr, prior = pr, cores = 4, sample_prior = "only")
```

```{r}
pp_check(mod_pr)
```

This prior seems really tight but actually allows for pretty high counts.
Now we can run the model with data:

```{r}
mod_p <- update(mod_pr, sample_prior = "no", cores = 4)
```

## Model checks!
Statistics are printed by summary
```{r}
summary(mod_p)
```
Plot the variables to see the traceplots
```{r}
plot(mod_p)
```

Alternatively, plot the rank overlay for the chains
```{r}
mcmc_plot(mod_p, type = "rank_overlay")
```

Now we can look at how well the model predicted the data using posterior predictive checks:
```{r}
pp_check(mod_p)
```

There are other types of posterior predictive checks supported by pp_check, described further in the documentation.

To examine what the model estimates the effect of treatment to be upon worm count we can plot the predicted response for the different predictors.
```{r}
fixef(mod_p)
plot(conditional_effects(mod_p))
```

# Errors
Stan returns many more potential errors and warnings than other MCMC software, in part because the fine-tuning of the NUTS algorithm offers more opportunities to pick up on issues with the exploration of model space. A full description of the different warnings is at https://mc-stan.org/misc/warnings.html but here's a quick summary of the ones I've commonly run into:

* divergent transitions - the warning message will recommend increasing adapt_delta which may work, if not then the model structure needs to change
* maximum treedepth exceeded - the warning message will recommend increasing max_treedepth (this is an efficiency concern, not a validity concern)
* Rhat - will return a warning if above 1.05. Note that stan now uses a more robust rhat so this will pick up on issues where the old version may not have.
* Effective sample size warnings for the bulk and tail of the distribution, will suggest running for more iterations but I've mostly run across these when chains haven't fully converged so fix that first  

# More complicated models
The above are quite simple examples, but brms can support many other types of model including those with missing data, censoring, multiple responses or non-linear models.


## Multivariate models
Modelling multiple response variables within brms can be done in one of two ways, if you have both response variables being predicted by the same predictors and having the same family you can use mvbind() to combine the two. Otherwise, you have to specify each formula within a bf() function then combine them together in the brm call. Fitting multiple models together allows you to model correlation between response variables and use information criteria or cross-validation upon the entire model.

```{r multi model fit}
dat <- australia.soybean %>%
  mutate(YR = as.factor(year)) %>%
  mutate_if(is.numeric, scale) %>%
  na.omit()

pr <- c(prior(normal(0,1), class = "b", resp = "protein"),
        prior(normal(0,1), class = "b", resp = "oil"))

mod_mv <- brm(mvbind(protein,oil) ~ year*loc, 
              data = dat, prior = pr, cores = 4)
summary(mod_mv)
```

```{r multi model loo}
plot(conditional_effects(mod_mv, effects = "year:loc", resp = "protein"))
plot(conditional_effects(mod_mv, effects = "year:loc", resp = "oil"))

mod_mv <- add_criterion(mod_mv, "loo")
print(mod_mv$criteria$loo)
```
Alternatively, if the two response variables have differing 


## Missing data
Missing data can be imputed using the mi() notation, you have to specify which predictors you want the model to use in imputing the missing data. This example is lifted directly from the missing data vignette in brms.

```{r missing data}
data("nhanes", package = "mice")
bform <- bf(bmi | mi() ~ age * mi(chl)) +
  bf(chl | mi() ~ age) + set_rescor(FALSE)
fit <- brm(bform, data = nhanes, cores = 4)
summary(fit)
plot(conditional_effects(fit, resp = "bmi"), ask = FALSE)
```




## Non-linear models
Non-linear models can also be fit within bf(), you have to specify that the model is non-linear (with nl = TRUE), and also specify the model parameters explicitly. If the model parameters are not dependent upon anything this takes the form of a param ~ 1 section, otherwise it can be a param ~ Variable section. The below example is based upon the example in the non-linear vignette in brms. 

```{r nonlinear data sim}
set.seed(45276)
b <- c(2, 0.75)
x <- rnorm(100)
y <- rnorm(100, mean = b[1] * exp(b[2] * x))
site <- gl(25,4)
dat1 <- data.frame(x, y, site)
```

```{r nonlinear model fit}
prior1 <- prior(normal(1, 2), nlpar = "b1") +
  prior(normal(0, 2), nlpar = "b2")
fit1 <- brm(bf(y ~ b1 * exp(b2 * x), b1 ~ (1|site), b2 ~ 1, nl = TRUE),
            data = dat1, prior = prior1, cores = 4,
            control = list(adapt_delta = 0.9))
summary(fit1)
```

```{r nonlinear model plots}
plot(fit1)
plot(conditional_effects(fit1), points = TRUE)
```

